---
title: "Introduction to QAIDR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to QAIDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

## Overview

**QAIDR** (Quality Assessment for Interval-based Dimensionality Reduction)
provides tools for evaluating interval DR methods using co-ranking matrices.
It supports:

- **4 interval distance metrics**: Interval Euclidean, Hausdorff,
  Ichino-Yaguchi, L2-Wasserstein
- **3 quality/behavior index families**: T&C, MRRE, LCMC (6 indices total)
- **6 DR methods**: C-PCA, V-PCA, MR-PCA, SPCA, IMDS, Int-UMAP
- **Statistical tests**: Permutation-based significance testing

## Quick Start

### 1. Creating Interval Data

Interval-valued data consists of observations where each variable is an
interval $[a, b]$, represented internally by its center $(a+b)/2$ and
radius $(b-a)/2$.

```{r create-data}
library(QAIDR)

# From centers and radii matrices
C <- matrix(rnorm(40), 10, 4)
R <- matrix(runif(40, 0.1, 0.5), 10, 4)
x <- interval_data(C, R, labels = rep(c("A", "B"), each = 5))
print(x)
```

You can also create interval data from common formats:

```{r from-mm}
# From a min-max data frame
df <- data.frame(
  A.min = c(1, 3, 5), A.max = c(2, 5, 8),
  B.min = c(10, 20, 30), B.max = c(12, 25, 35),
  class = c("x", "y", "x")
)
x2 <- interval_data_from_mm(df, int_min_at = c(1, 3), y_at = 5)
print(x2)

# From a 3D array (n x p x 2)
arr <- array(c(1, 2, 3, 4, 5, 6, 2, 4, 5, 6, 8, 9), dim = c(3, 2, 2))
x3 <- interval_data_from_array(arr)
```

### 2. Standardization

Standardize interval data so that each variable has zero-mean centers and
unit standard deviation:

```{r standardize}
xs <- standardize(x)
# Column means of centers are now 0
round(colMeans(xs$centers), 10)
# Column SDs of centers are now 1
round(apply(xs$centers, 2, sd), 10)
```

### 3. Computing Interval Distances

Four distance metrics are available for comparing interval-valued
observations:

```{r distances}
D1 <- idist(xs, metric = "Wasserstein")
D2 <- idist(xs, metric = "Hausdorff")
D3 <- idist(xs, metric = "Int-Euclidean")
D4 <- idist(xs, metric = "Ichino-Yaguchi")

cat("Distance matrix dimensions:", dim(D1), "\n")
```

### 4. Co-ranking Indices

Given high-dimensional and low-dimensional distance matrices, compute the
six quality/behavior indices:

```{r coranking}
set.seed(42)
Y <- matrix(rnorm(20), 10, 2)
Dl <- as.matrix(dist(Y))
Dh <- idist(xs, metric = "Wasserstein")

indices <- coranking_indices(Dh, Dl, K = 3)
print(indices)
```

- **Quality indices** (Q_TC, Q_RE, Q_LC): range [0, 1], higher is better
- **Behavior indices** (B_TC, B_RE, B_LC): range [-1, 1], 0 means balanced

### 5. Using Built-in Datasets

```{r datasets}
data(cars_mm)
print(cars_mm)
summary(cars_mm)

data(facedata_mm)
print(facedata_mm)
```

## Full Workflow

The typical QAIDR workflow is:

1. Load or create `interval_data`
2. Standardize with `standardize()`
3. Run DR methods with `run_idr()`
4. Assess quality with `assess_quality()`
5. Visualize with `plot_projections()` and `plot_k_profiles()`

```{r full-workflow, eval = FALSE}
# Complete example (requires symbolicDA, RSDA)
data(cars_mm)
x <- standardize(cars_mm)

proj <- run_idr(x)
result <- assess_quality(x, proj, K = 5, perm_test = TRUE, n_perm = 1000)
print(result)

plot_projections(proj, labels = cars_mm$labels)

profiles <- k_profiles(x, proj, K_max = 20)
plot_k_profiles(profiles, metric = "Wasserstein")
```

See the "Real Data Analysis" and "Simulation Study" vignettes for complete
examples.
