---
title: "Real Data Analysis with QAIDR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real Data Analysis with QAIDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  eval = FALSE
)
```

This vignette reproduces the real data analysis from the QAIDR paper, applying
six interval dimensionality reduction methods to the Cars and Face datasets and
evaluating them with co-ranking quality and behavior indices.

**Prerequisites:** The DR methods require `symbolicDA`, `RSDA`, and `umap`
packages. Install them with:

```{r install, eval = FALSE}
install.packages(c("symbolicDA", "RSDA", "umap"))
```

## Cars Dataset

The Cars dataset contains 27 car models described by 4 interval-valued
variables (Price, Engine Capacity, Top Speed, Acceleration) with 4 class
labels (Berlina, Luxury, Sportive, Utilitarian).

```{r cars-setup}
library(QAIDR)
set.seed(2025)

data(cars_mm)
print(cars_mm)
summary(cars_mm)
```

### Standardize

```{r cars-std}
x <- standardize(cars_mm)
```

### Run all 6 DR methods

```{r cars-dr}
proj <- run_idr(x, labels = cars_mm$labels)
print(proj)
```

### 2D projection plots

Each DR method produces a 2D projection where intervals are displayed as
rectangles:

```{r cars-proj, fig.width = 12, fig.height = 8}
plot_projections(proj,
                 labels = cars_mm$labels,
                 obs_labels = rownames(cars_mm$centers))
```

### Quality assessment with permutation tests

Evaluate all 6 methods across 4 metrics at neighbourhood size K = 5, with
1000 permutations for significance testing:

```{r cars-assess}
result <- assess_quality(x, proj, K = 5,
                         perm_test = TRUE, n_perm = 1000)
print(result)
```

An asterisk (`*`) indicates statistical significance at the 0.05 level.

### K-neighbourhood profiles

Quality and behavior indices across all neighbourhood sizes K:

```{r cars-profiles, fig.width = 12, fig.height = 9}
profiles <- k_profiles(x, proj)

# Plot for each metric
for (met in c("Int-Euclidean", "Hausdorff", "Ichino-Yaguchi", "Wasserstein")) {
  plot_k_profiles(profiles, metric = met)
}
```

## Face Dataset

The Face dataset contains 27 individuals described by 6 interval-valued
anthropometric measurements (AD, BC, AH, DH, EH, GH).

```{r face-setup}
data(facedata_mm)
print(facedata_mm)
```

### Standardize and run DR

```{r face-dr}
x_face <- standardize(facedata_mm)
proj_face <- run_idr(x_face, labels = facedata_mm$labels)
```

### Projection plots

```{r face-proj, fig.width = 12, fig.height = 8}
plot_projections(proj_face,
                 labels = facedata_mm$labels,
                 obs_labels = rownames(facedata_mm$centers))
```

### Quality assessment

```{r face-assess}
result_face <- assess_quality(x_face, proj_face, K = 5,
                              perm_test = TRUE, n_perm = 1000)
print(result_face)
```

### K-neighbourhood profiles

```{r face-profiles, fig.width = 12, fig.height = 9}
profiles_face <- k_profiles(x_face, proj_face)
plot_k_profiles(profiles_face, metric = "Wasserstein")
```

## Interpreting Results

The assessment table reports six indices for each method-metric combination:

| Index | Type | Range | Interpretation |
|-------|------|-------|----------------|
| Q_TC | Quality | [0, 1] | Average of Trustworthiness and Continuity |
| B_TC | Behavior | [-1, 1] | Continuity - Trustworthiness (+ = extrusion-dominant) |
| Q_RE | Quality | [0, 1] | 1 - average relative rank error |
| B_RE | Behavior | [-1, 1] | Intrusion error - Extrusion error |
| Q_LC | Quality | [0, 1] | Fraction of K-neighbours preserved |
| B_LC | Behavior | [-1, 1] | Asymmetry in local continuity |

**Quality indices** closer to 1 indicate better structure preservation.
**Behavior indices** near 0 indicate balanced intrusions and extrusions;
large positive or negative values indicate directional bias in the
embedding.
