---
title: "Real-Data Examples with QAIDR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-Data Examples with QAIDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  eval = FALSE
)
```

This vignette demonstrates the QAIDR workflow with the `cars.int` and
`face.iGAP` datasets distributed by the CRAN package `dataSDA`. It is a
package tutorial, not an exact reproduction of the manuscript. The paper's
real-data study uses Face and a 645-station USHCN dataset; its seeded scripts,
frozen processed data, and outputs are documented in the GitHub
`README_REPRODUCIBILITY.md` file.

The DR wrappers require `symbolicDA`; Int-UMAP additionally requires `RSDA`,
`umap`, and `dplyr`.

```{r install}
install.packages(c("dataSDA", "symbolicDA", "RSDA", "umap", "dplyr"))
```

## Cars data

`cars.int` contains 27 car models, four interval-valued variables (Price,
Engine Capacity, Top Speed, and Acceleration), and a class variable. The
conversion keeps the model names stored in the data's `concept` attribute and
uses the singleton symbolic sets in `class` as labels.

```{r cars-setup}
library(QAIDR)
data("cars.int", package = "dataSDA")

cars <- interval_data_from_dataSDA(cars.int)
print(cars)
summary(cars)
x_cars <- standardize(cars)
```

Run the available DR methods and inspect the projected rectangles:

```{r cars-dr}
set.seed(20260718)
proj_cars <- run_idr(x_cars)
plot_projections(proj_cars, labels = cars$labels,
                 obs_labels = rownames(cars$centers))
```

Assess every method--dissimilarity pair at `K = 5`. The center-only row is an
essential comparison: it shows whether interval widths add assessment
information beyond the midpoints alone. The default 999 permutations permit
finite-sample p-values below 0.05, and `pvalues_adj` contains the within-family
Westfall--Young min-P adjustment.

```{r cars-assess}
cars_assessment <- assess_quality(
  x_cars, proj_cars, K = 5,
  perm_test = TRUE, n_perm = 999,
  seed = 20260719, baseline = TRUE
)
print(cars_assessment)
cars_assessment$pvalues_adj
```

Quality and behavior can also be inspected across neighborhood sizes:

```{r cars-profiles}
cars_profiles <- k_profiles(x_cars, proj_cars, K_max = 20, baseline = TRUE)
plot_k_profiles(cars_profiles, metric = "Wasserstein")
```

## Face data

`face.iGAP` contains six interval-valued anthropometric measurements for
27 records. Its row names identify nine people with three replicates each; we
derive the person labels by removing the trailing replicate number.

```{r face-setup}
data("face.iGAP", package = "dataSDA")
face_labels <- sub("[[:digit:]]+$", "", rownames(face.iGAP))
face <- interval_data_from_dataSDA(face.iGAP, labels = face_labels)
x_face <- standardize(face)
print(face)
```

```{r face-analysis}
set.seed(20260720)
proj_face <- run_idr(x_face)
plot_projections(proj_face, labels = face$labels,
                 obs_labels = rownames(face$centers))

face_assessment <- assess_quality(
  x_face, proj_face, K = 5,
  perm_test = TRUE, n_perm = 999,
  seed = 20260721, baseline = TRUE
)
print(face_assessment)

face_profiles <- k_profiles(x_face, proj_face, K_max = 20, baseline = TRUE)
plot_k_profiles(face_profiles, metric = "Wasserstein")
```

## Interpreting the indices

| Index | Type | Interpretation |
|:------|:-----|:---------------|
| Q_TC | Quality | Average of trustworthiness and continuity |
| B_TC | Behavior | Continuity minus trustworthiness |
| Q_RE | Quality | One minus the average relative rank error |
| B_RE | Behavior | Intrusion error minus extrusion error |
| Q_LC | Quality | Fraction of K-neighbors preserved |
| B_LC | Behavior | Bounded local intrusion--extrusion asymmetry |

Higher quality values indicate better neighborhood preservation. Behavior
values near zero indicate balance; positive values consistently indicate an
intrusion-dominated embedding, and negative values indicate an
extrusion-dominated embedding. Interpret isolated values together with their
K-profiles, tie diagnostics, center-only baseline, and adjusted permutation
results.
