---
title: "Simulation Study with QAIDR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study with QAIDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  eval = FALSE
)
```

This vignette reproduces the simulation study from the QAIDR paper. Three
scenarios test different aspects of the quality and behavior indices.

**Prerequisites:** Requires `symbolicDA`, `RSDA`, and `umap` packages.

## Common Setup

```{r common-setup}
library(QAIDR)
set.seed(2025)

methods_list <- c("C-PCA", "V-PCA", "MR-PCA", "SPCA", "IMDS", "Int-UMAP")
metrics_list <- c("Int-Euclidean", "Hausdorff", "Ichino-Yaguchi", "Wasserstein")

# Set N_PERM = 1000 for final paper results
N_PERM <- 10
```

## Scenario I: Varying Variability

Tests how the indices respond to different radii magnitudes. Three groups of
100 observations each have small, medium, and large interval widths.

```{r scenario1-data}
N <- 300; P <- 5
centers <- matrix(rnorm(N * P), N, P)
radii <- matrix(0, N, P)

# Group 1: Small radii (0.1 -- 0.2)
radii[1:100, ] <- matrix(runif(100 * P, 0.1, 0.2), 100, P)
# Group 2: Medium radii (1.0 -- 1.2)
radii[101:200, ] <- matrix(runif(100 * P, 1.0, 1.2), 100, P)
# Group 3: Large radii (5.0 -- 6.0)
radii[201:300, ] <- matrix(runif(100 * P, 5.0, 6.0), 100, P)

colnames(centers) <- paste0("V", seq_len(P))
colnames(radii) <- paste0("V", seq_len(P))

x_s1 <- interval_data(centers, radii)
x_s1_std <- standardize(x_s1)
```

### Run DR and evaluate

```{r scenario1-run}
proj_s1 <- run_idr(x_s1_std)

res_s1 <- data.frame()
for (met in methods_list) {
  cat("Processing:", met, "\n")
  out <- proj_s1[[met]]

  for (metric in metrics_list) {
    Dh <- idist(x_s1_std, metric = metric)
    Dl <- if (out$type == "Point") {
      as.matrix(dist(out$C))
    } else {
      idist(out$C, out$R, metric)
    }

    pt <- perm_test(Dh, Dl, K = 10, n_perm = N_PERM)
    res_s1 <- rbind(res_s1, data.frame(
      Method = met, Metric = metric,
      Q_TC = sprintf("%.3f", pt$vals["Q_TC"]),
      B_TC = sprintf("%.3f", pt$vals["B_TC"]),
      Q_RE = sprintf("%.3f", pt$vals["Q_RE"]),
      B_RE = sprintf("%.3f", pt$vals["B_RE"]),
      Q_LC = sprintf("%.3f", pt$vals["Q_LC"]),
      B_LC = sprintf("%.3f", pt$vals["B_LC"])
    ))
  }
}
print(res_s1)
```

Or more concisely with `assess_quality()`:

```{r scenario1-concise}
result_s1 <- assess_quality(x_s1_std, proj_s1, K = 10,
                            perm_test = TRUE, n_perm = N_PERM)
print(result_s1)
```

## Scenario II: Interval Swiss Roll

Tests how well DR methods preserve the structure of a nonlinear manifold
when data are interval-valued.

```{r scenario2-data}
dat_sr <- gen_interval_swissroll(n = 800, seed = 42)
print(dat_sr)
```

### Standardize and run DR

```{r scenario2-run}
dat_sr_std <- standardize(dat_sr)
proj_sr <- run_idr(dat_sr_std)

res_s2 <- data.frame()
for (met in methods_list) {
  cat("Processing:", met, "\n")
  out <- proj_sr[[met]]

  for (metric in metrics_list) {
    Dh <- idist(dat_sr_std, metric = metric)
    Dl <- if (out$type == "Point") {
      as.matrix(dist(out$C))
    } else {
      idist(out$C, out$R, metric)
    }

    pt <- perm_test(Dh, Dl, K = 10, n_perm = N_PERM)
    res_s2 <- rbind(res_s2, data.frame(
      Method = met, Metric = metric,
      Q_TC = sprintf("%.3f", pt$vals["Q_TC"]),
      B_TC = sprintf("%.3f", pt$vals["B_TC"]),
      Q_RE = sprintf("%.3f", pt$vals["Q_RE"]),
      B_RE = sprintf("%.3f", pt$vals["B_RE"]),
      Q_LC = sprintf("%.3f", pt$vals["Q_LC"]),
      B_LC = sprintf("%.3f", pt$vals["B_LC"])
    ))
  }
}
print(res_s2)
```

## Scenario III: Stability Analysis

Tests robustness of indices under small perturbations of interval radii.
Mean Absolute Error (MAE) and sign-flip rates are computed over 100
iterations.

```{r scenario3-setup}
N <- 200; P <- 5; K <- 10; N_ITER <- 100
center_sd <- 0.05; r_min <- 1.2; r_max <- 2.5; eps <- 0.25

# Generate base data with dense overlap
centers_base <- matrix(rnorm(N * P, mean = 0, sd = center_sd), N, P)
radii_base <- matrix(runif(N * P, r_min, r_max), N, P)
colnames(centers_base) <- paste0("V", seq_len(P))
colnames(radii_base) <- paste0("V", seq_len(P))

x_base <- interval_data(centers_base, radii_base)
xs_base <- standardize(x_base)
proj_base <- run_idr(xs_base)
```

### Compute MAE and sign-flip rates

```{r scenario3-run}
stability <- data.frame()

for (met in names(proj_base)) {
  for (mn in metrics_list) {
    # Baseline indices
    Dh0 <- idist(xs_base, metric = mn)
    out <- proj_base[[met]]
    Dl <- if (out$type == "Point") {
      as.matrix(dist(out$C))
    } else {
      idist(out$C, out$R, mn)
    }
    base_vals <- coranking_indices(Dh0, Dl, K = K)

    # Perturb radii N_ITER times
    vals <- matrix(NA, nrow = N_ITER, ncol = 6)
    for (i in seq_len(N_ITER)) {
      mult_noise <- matrix(runif(N * P, -eps, eps), N, P)
      r_pert <- radii_base * (1 + mult_noise)
      r_pert[r_pert < 1e-8] <- 1e-8
      Dh <- idist(centers_base, r_pert, mn)
      vals[i, ] <- coranking_indices(Dh, Dl, K = K)
    }

    # MAE
    mae <- colMeans(abs(vals - matrix(base_vals, nrow = N_ITER,
                                      ncol = 6, byrow = TRUE)))

    # Sign-flip rate for behavior indices (columns 2, 4, 6)
    sign_flip <- c(
      mean(sign(vals[, 2]) != sign(base_vals[2])),
      mean(sign(vals[, 4]) != sign(base_vals[4])),
      mean(sign(vals[, 6]) != sign(base_vals[6]))
    )

    stability <- rbind(stability, data.frame(
      Method = met, Metric = mn,
      MAE_Q_TC = mae[1], MAE_B_TC = mae[2],
      MAE_Q_RE = mae[3], MAE_B_RE = mae[4],
      MAE_Q_LC = mae[5], MAE_B_LC = mae[6],
      SignFlip_B_TC = sign_flip[1],
      SignFlip_B_RE = sign_flip[2],
      SignFlip_B_LC = sign_flip[3]
    ))
  }
}

print(stability)
```

## Notes

- For reproducible paper results, set `N_PERM = 1000` and `N_ITER = 100`.
- The `eval = FALSE` default avoids long computation during vignette building.
  Copy the code into an R session with the required packages installed to
  run interactively.
- Scenario III is designed with dense overlap (`center_sd = 0.05`,
  `r_min = 1.2`) to stress-test the stability of the indices.
